Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-06-28 MIMOSA.tex
Results and discussion
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[1]{Raphael Gottardo}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, Wa}
\date{\today}

@@

\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see Supplementary Information~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

@@

We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 98 individuals. We examined time points from day 0 and day 182 (the primary endpoint).We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples. An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.

\subsection{Fluidigm}
We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. The unconstrained MIMOSA model was fit separately to for each stimulation, and the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions were calculated and heat maps generated (Figure~\ref{fit:fluidigm} A-C).

\subsection{Simulation Studies}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the model fit of the constrained model to data from Gag1 stimulated CD4--positive, IL2 expressing T--cells on day 28 from real data, we simulated data from the constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, and ten independent realizations for each $N$. The constrained model was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through ROC curve analysis and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change. (Figure~\ref{fig:simulations}). The nominal vs observed false discovery rate was also examined to assess the model fit.

For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, B). The false discovery rate for MIMOSA more closely reflected the nominal false discovery rate than Fisher's exact test (Figure~\ref{fig:simulations} C, D ).
%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the MLE estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained model fit to constrained data.

@@

\section{Discussion}
The variety of single--cell assays being adopted by the immunology community is increasing. Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data. Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}. Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.

The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model cell counts in non--responding and responding individuals. Information is shared across non--responders and across responders through an empirical--Bayes prior, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}). The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample, our simulations show that we can relax this assumption to the unconstrained case where $p_s\ne p_u$ without compromising the observed false positive or false negative rates (Figures~\ref{fig:simulations} and \ref{fig:simulations_trunc}), while saving on computation time.

\todo[inline]{Fix references to non-existent figure}
The analysis of real--world ICS data from a vaccine trial demonstrated that the MIMOSA model outperformed Fisher's exact test in detecting individuals who were responders and non--responders to vaccine across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:positivityrates}). Importantly, the MIMOSA model was fit within each antigen stimulation $\times$ cytokine $\times$ time point combination but the model was blinded to placebo and vaccine treatment. Despite this, MIMOSA demonstrated a lower false positive rate at day zero (increased specificity), where no response is expected, than Fisher's exact test (Figure~\ref{fig:positivityrates} Gag2, IL2, CD4+ T--cells, day 0, treatment group T1). Similarly, none of the placebo treated samples were identified as responders by the MIMOSA model (Figure~\ref{fig:positivityrates} treatment group C). MIMOSA also exhibited an increased sensitivity in real--world data compared to Fisher's exact test, as demonstrated by the higher response rates for 7 of the 8 stimulation/cytokine/cell type combinations at day 28 to which the model was fit, and where a response is expected to be observed. Although it is not possible to validate the responders identified at day 28 for all cases, the conclusion of increased sensitivity is consistent with both the performance of the model in simulations, and with the increased specificity observed on day 0 and in the placebo treated samples.

\todo[inline]{Fix references to non--existent figure}
%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).  For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test. These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions. This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model. This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.

Polyfunctionality (identifying cells coexpressing multiple proteins, cytokines, or genes, is important in single cells assays since it allows the identification of functionally distinct subsets of cells in heterogeneous cell populations~\cite{Milush:2009bz}. In vaccine trials, these polyfunctional cell subsets may be correlated with vaccine response or efficacy. Existing methods for identifying polyfunctional cytokine profiles from ICS data either test individual combinations of cytokines separately, perform global tests at the expense of decreased power when only specific differences are of interest, sometimes  empirical in nature, and generally don't share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}. As others have pointed out, in order to have the most power to detect a true difference, the statistical test selected to detect differences in various combinations of cytokines should take into consideration the cytokine combinations that are of interest~\cite{Nason:2006dx}. A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different. This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg. ROC and FDR plots of other cytokine combinations can be found in the Supplementary Information.}
\label{fig:HVTN065}
\end{figure}

@@

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row). Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data. A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change. B) Boxplots for the unconstrained model. C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test. C) ROC curves for a representative simulated data set (N=100,000 counts). D) AUC  boxplots, E) observed vs. nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data. }\label{fig:simulations}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
    \node at (0,4.2) [font=\tiny\sffamily] {A} ;
    \node at (4.6,4.2) [font=\tiny\sffamily] {B} ;
    \node at (9,4.2) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
   \caption{Multivariate simulations from a two--sided model. Ten, five-dimensional data sets were simulated from a two--sided model with an effect size of 2.5e-3 and -2.5e-3 in two of the five dimensions. Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions. A) Boxplots of AUCs (area under the curve) for the ten simulations. B) Observed and nominal false discovery rate for each method. C)ROC curves for the competing methods.}
   \label{fig:mvsimulations}
\end{figure}

>>>>

\title{Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[3]{Mario Roederer}
\author[1]{Raphael Gottardo}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\affil[3]{Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}

\date{\today}

@@

\textit{Flow cytometry}: This dataset is from a phase-I (safety and efficacy) trial of an adenoviral vector vaccine in individuals without prior immunity, measuring four cytokines via intracellular cytokine staining (ICS) in two cell populations from 20 individuals at two time points (zero and 28 days post-vaccination, see supplementary material~\ref{supp:data} for details)~\cite{Peiperl:2010ej}. The statistical analysis of responder\footnote{need to define this} and non-responder calls in the published trial is described in the original manuscript and outlined in the supplementary information (Supplementary Information~\ref{supp:statpublished})~\cite{Peiperl:2010ej}. The goal of this data set was to assess and quantify response rates of CD4 and CD8 T-cell populations to different antigens.

@@

We tested our method on ICS data from HVTN065,  a trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen, with 120 individuals (98 vaccines and 20 placebo recipients). We analyzed the vaccinees (n=98) at the day 0 and day 182 (the primary endpoint) time points as a pooled data set. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity). We examined the CD4+ T--cell cytokine responses for ENV-1-PTEG and GAG-1-PTEG stimulated samples. An ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the constrained MIMOSA model compared to Fisher's exact test, ranked log fold change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non--responders.

The MMOSA model had higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or ranked log fold change for discriminating vaccine responders and non--responders when the magnitude of response was small (i.e. ENV-1-PTEG stimulation) (Figure~\ref{fig:HVTN065} A),  (AUC=0.954 for MIMOSA vs. 0.923 for Fisher's exact test vs. 0.827 for LRT vs. 0.775 for ranked fold change). When the magnitude of response was larger (GAG-1-PTEG stimulation), MIMOSA performed as well as Fisher's exact test, and better than other competing methods (Figure~\ref{fig:HVTN065} C) (AUC=0.982 for MIMOSA vs. 0.978 for Fisher's exact test vs. 0.877 for LRT vs. 0.784 for ranked fold change). Irrespective of the magnitude of the response to stimulation, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B,D). These results are consistent for other cytokines (see supplementary material).

\subsection{Fluidigm}
\subsubsection{MIMOSA Identifies Stimulation-Specific Patterns of Gene Expression in Fluidigm Single--Cell Data}
We applied the MIMOSA model to Fluidigm single--cell gene expression data from CD8+ T--cells from 16 individuals, under four different stimulation conditions, as well as unstimulated samples. The unconstrained MIMOSA model was fit to each stimulation. From the model we calculated the posterior probabilities of response for each gene, the posterior differences in the proportion of cells expressing each gene between stimulated and unstimulated conditions, as well as the posterior log ratio of the proportion of cells expressing each gene in stimulated and unstimulated conditions. The data are presented in (Figure~\ref{fig:fluidigm} A-C). Importantly, we see that MIMOSA identifies stimulation--specific differences in the proportions of cells expressing each gene, while preserving inter--subject variability (Figure~\ref{fig:fluidigm}). These patterns are preserved in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), as well as in the posterior differences and log--odds (Figure~\ref{fig:fluidigm} B,C).

\subsection{Simulation Studies}
We examined the performance of the constrained ($p_s>p_u$) and unconstrained ($p_s \ne p_u$) beta--binomial mixture models via simulations. Using hyper parameters estimated from the constrained model fit to data from Gag1--stimulated, CD4--positive, IL2--expressing T--cells from day 28 post--vaccination of the HVTN054 trial~\cite{Peiperl:2010ej}. We simulated data from this constrained model with 500 observations, a response rate of 60\%, an $N$ of 10K, 20K, 30K, 50K, 75K, 100K, and 150K events, with ten independent realizations of data for each $N$. Constrained MIMOSA was fit to this data and the sensitivity and specificity of the model's ability to correctly identify observations from the ``responder'' and ``non--responder'' components was evaluated through analysis of ROC curves, and compared against Fisher's exact test, the likelihood ratio test, and ranked log fold change. This procedure was repeated for the unconstrained model fit to unconstrained data (Figure~\ref{fig:simulations} A-D). The nominal vs observed false discovery rate was also examined to assess the model fit (Figure~\ref{fig:simulations} E-F).

For both the constrained and unconstrained simulations, MIMOSA out--performed competing methods, including Fisher's exact test, with respect to sensitivity and specificity at all values of $N$ (Figure~\ref{fig:simulations} A, D). Additionally, the false discovery rate for MIMOSA more closely reflected the nominal false discovery rate compared to Fisher's exact test (Figure~\ref{fig:simulations} E, F).
%Since the constrained model relies on Monte--Carlo integration to estimate the normalizing constant in the likelihood calculation, which can be computationally costly, we approximated the.

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

@@

\subsubsection{Polyfunctionality in Fluidigm Single Cell Gene Expression Data}
We fit the multivariate MIMOSA model to the two--sided Fluidigm data, looking at coexpression of pairs of genes in single cells across the four stimulations. In Figure~\ref{fig:polyfunctionality} we show heatmaps of the counts of cells expressing all combinations of the BIRC3 and CCL5 genes in unstimulated and stimulated samples (Figure~\ref{fig:polyfunctionality} A,B). Only CCL5 positive cells express BIRC3, and the expression of BIRC3 increases upon stimulation. The typical approach to analyzing polyfunctional populations from intracellular cytokine staining data (summing the counts over all possible polyfunctional cell populations) would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. That is to say, the number of BIRC3-/CCL5+ cells decreases upon stimulation and the number of BIRC3+/CCL5+ cells increases. There is no difference in the marginal counts for any sample (Figure~\ref{fig:polyfunctionality} C). In contrast, multivariate MIMOSA tests all cell subpopulations simultaneously, and identifies a significant difference between stimulated and unstimulated conditions in 13 of the 16 samples (Figure~\ref{fig:polyfunctionality} D). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the individual combinations, which requires multiplicity adjustment and a potential loss of power.

Since the Fluidigm data has a limited number of observations (100 cells and 16 samples), we performed simulations (five--dimensional data) to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test on the resulting 2x5 tables, Fisher's exact test performed marginally on each dimension of the data and p--values combined through Fisher's method ($\chi^2=-2\sum_{i=1}^k\log(p_i)$), as well as the likelihood ratio test (Figure~\ref{fig:mvsimulations} A-C). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model is a better fit to the data than other standard approaches tested for analyzing such multivariate count data.

\section{Discussion}
The variety of single--cell assays being adopted by the immunology community is increasing. Flow cytometry, mass cytometry, ELISPOT, Fluidigm, and other single--cell assays can all be analyzed as single--cell count data. Development of effective statistical methods to detect differences in gene or protein expression at the single cell level is becoming increasingly important. Current approaches rely on asymptotic approximations (t--test, or $\chi^2$ test), empirical or ad--hoc methods (2--fold change), or exact tests (Fisher's exact test) where model assumptions are generally not satisfied, all of which can lead to invalid conclusions about the data~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks}. Most importantly, existing classical methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modelling approaches, which are widely applied in the microarray literature~\cite{Kendziorski:2003uw,Newton:2001go,Smyth:2005iy}.

The MIMOSA model presented here uses a mixture model framework of Beta--Binomial or Multinomial--Dirichlet distributions to model counts in experimental units (i.e. individuals) across multiple conditions (i.e. vaccine responders and non--responders). Information is shared across non--responder individuals and across responders individuals through exchangeable Beta or Dirichlet priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when model assumptions are violated (Figures~\ref{fig:simulations} and~\ref{fig:simulations_trunc}). The MIMOSA model based on the Beta--Binomial distribution allows us to constrain the alternative hypothesis to the case $p_s > p_u$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample.

Importantly, the analysis of real--world ICS data from vaccine trials demonstrated that the the constrained MIMOSA model performs as well or better than Fisher's exact test or other non-Bayesian alternatives for identifying vaccine responders across multiple antigen stimulations and multiple cytokines (Figure~\ref{fig:HVTN065}). Although the MIMOSA model was fit within each antigen stimulation $\times$ cytokine combination but the model was blinded to vaccine time point. Despite this, MIMOSA demonstrated a higher sensitivity and specificity to discriminate between vaccine responders and non-responders on days 0 (pre--vaccination) and 182 (post--vaccination) than the current standard approach (Fisher's exact test), or non--Bayesian alternatives such as the likelihood ratio test or ranking by log--fold change (Figure~\ref{fig:HVTN065} A,C). Our approach treats all day 0 observations as true negatives and all day 182 observations as true positives, yet we know that not all vaccine recipients are likely to exhibit a vaccine response on day 182. None the less, this shortcoming, at worst, leads us to underestimate the true sensitivity of MIMOSA, and does not negatively impact the comparison. Additionally, MIMOSA was found to be a better fit to real--world ICS data from vaccine trials than other analysis approaches, as evidenced by more accurate estimates of true false discovery rate (Figure~\ref{fig:HVTN065} B,D).

%We examined some of the contradictory observations identified as responders by MIMOSA and non--responders by Fisher's exact test more closely (Figure~\ref{fig:icsdata}).  For Gag2 stimulated CD4+ T--cells expressing IFNg, we see two observations detected as responders by MIMOSA at the 1\% FDR level, but not by Fisher's exact test. These two observations had a 2.32 and 3.08 fold increase in the proportion of cytokine--positive cells in the stimulated vs unstimulated samples, respectively, but differences in absolute positive cell counts of 2-3 cells, although the negative cell counts for the simulated and unstimulated samples were substantially different (32K vs 82K cells, and 65K vs 145K cells, respectively), resulting in the significant difference in proportions. This is a clear example where sharing information through the empirical Bayes framework, and eliminating the conditioning on fixed margins inherent in Fisher's exact test, lead to improved sensitivity on real data for the MIMOSA model. This increased power to detect vaccine responders is important since it impacts decision--making about which vaccine candidates move forward to later phase clinical trials.

Measuring differences in polyfunctional cell populations (i.e. identifying changes in cell populations that co--express multiple proteins, cytokines, or genes), is important in single cell immunological assays since they allow the identification of functionally distinct cell subsets in heterogeneous populations~\cite{Milush:2009bz}. In vaccine trials, these polyfunctional cell subsets may be correlated with differential vaccine response or efficacy and have the potential to inform vaccine development and trial design. Existing methods for identifying polyfunctional cytokine profiles from ICS data either separately test individual combinations of cytokines or perform global tests between groups of individuals even when the populations under study are known to be heterogeneous (i.e. mixtures of responders and non--responders).  Sometimes tests are empirical in nature (i.e. thresholds based on fold change or some factor over background), and these methods generally do not share information across observations~\cite{Dittrich:2012bv,Trigona:2003,Sinclair:2004hs,Horton:2007tsa,Proschan:2009ks,Nason:2006dx}. As others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account the cytokine combinations of interest~\cite{Nason:2006dx}. A global test, one that tests for a difference between all proportions of cytokine expressing cells will have lower power than one testing for difference in a specific cytokine combination if only that specific combination is truly different. This approach is inherent in our mixture model framework, where each mixture component representing some pattern of cytokine expression that is of interest.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=.75\columnwidth]{Figures/2}\\ 
%   \includegraphics[width=.75\columnwidth]{Figures/12} 
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \node[anchor=south west, inner sep=0] at (0,-3.75){\includegraphics[width=.6\columnwidth]{Figures/12}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {C} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {D} ;
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) GAG-1-PTEG and expressing IL2 and IFNg. ROC and FDR plots of other cytokine combinations can be found in the Supplementary material.}
\label{fig:HVTN065}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row). Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data. A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change. B) Boxplots for the unconstrained model. C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test. C) ROC curves for a representative simulated data set (N=100,000 counts). D) AUC  boxplots, E) observed vs. nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data. }\label{fig:simulations}
\end{figure}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/Multivariate_Sims_AUCBOX.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Multivariate_Sims_ROC.pdf}};
    \node at (0,2.1) [font=\tiny\sffamily] {A} ;
    \node at (4.5,2.1) [font=\tiny\sffamily] {B} ;
    \node at (9,2.1) [font=\tiny\sffamily] {C} ;
\end{tikzpicture}
   \caption{Multivariate simulations from a two--sided model. Ten, five-dimensional data sets were simulated from a two--sided model with an effect sizes of $2.5\times 10^{-3}$ and $-2.5\times 10^{-3}$ in two of the five dimensions. Multivariate MIMOSA was compared against Fisher's exact test, the likelihood ratio test, and Fisher's combined p--value, combining Fisher's exact test run marginally on each of the five dimensions. A) Boxplots of AUCs (area under the curve) for the ten simulations. B) Observed and nominal false discovery rate for each method. C)ROC curves for the competing methods.}
   \label{fig:mvsimulations}
\end{figure}
